---
title: "Ch02 R Codes"
author: "Ping-Yang Chen"
date: "2024-02-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Textbook: Montgomery, D. C. (2012). *Design and analysis of experiments*, 8th Edition. John Wiley & Sons.

# Chapter 2

Read Table 2.1
```{r, eval=TRUE, echo=TRUE}
df1 <- read.csv(file.path("data", "2_TensionBondStrength.csv"))
```


Draw grouped boxplot
```{r, eval=TRUE, echo=TRUE, fig.align='center'}
library(reshape2)
df1_L <- stack(df1, select = c("ModifiedMortar","UnmodifiedMortar"))
colnames(df1_L) <- c("Strength", "Mortar")
boxplot(Strength ~ Mortar, data = df1_L)
```

- `mean()`
- `median()`
- `min()` for the minimum,
- `max()` for the maximum,
- `sd()` for the standard deviation and
- `IQR()` for the interquartile range.
- The `summary()` function combines several of these statistics into one function call.

```{r, eval=TRUE, echo=TRUE, fig.align='center'}
sd(df1$ModifiedMortar)
sd(df1$UnmodifiedMortar)
```


```{r, eval=TRUE, echo=TRUE, fig.align='center'}
summary(df1)
```


```{r, eval=TRUE, echo=TRUE, fig.align='center'}
t.test(df1$ModifiedMortar, df1$UnmodifiedMortar,
       alternative = "two.sided",
       paired = FALSE, var.equal = TRUE)
```



## t-test unequal variance

Read Table 2.1
```{r, eval=TRUE, echo=TRUE}
df2 <- read.csv(file.path("data", "2_Fluorescence.csv"))
```


```{r, eval=TRUE, echo=TRUE, fig.align='center'}
sd(df2$Nerve)
sd(df2$Muscle)
```

```{r, eval=TRUE, echo=TRUE, fig.align='center'}
summary(df2)
```


```{r, eval=TRUE, echo=TRUE, fig.align='center'}
t.test(df2$Nerve, df2$Muscle,
       alternative = "two.sided",
       paired = FALSE, var.equal = FALSE)
```