---
title: "Ch-05 & 06 R Codes"
author: "Ping-Yang Chen"
#date: "2024-04-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Textbook: Montgomery, D. C. (2012). *Design and analysis of experiments*, 8th Edition. John Wiley & Sons.

Online handouts: https://github.com/PingYangChen/ANOVA_Course_R_Code

# Chapter 5: Factorial Experiments


```{r, eval=FALSE, echo=FALSE}
demo1 <- read.csv(file.path("data", "5_TwoFactorDesign.csv"))
demo2 <- read.csv(file.path("data", "5_TwoFactorDesignInteraction.csv"))
demo1$FactorA <- as.factor(demo1$FactorA)
demo1$FactorB <- as.factor(demo1$FactorB)
demo2$FactorA <- as.factor(demo2$FactorA)
demo2$FactorB <- as.factor(demo2$FactorB)
```

## Example 5.1 The Battery Life Experiment

Read the csv file `5_BatteryLife.csv` in R. Make sure that in the `data.frame` the variables `MaterialType` and `Temperature` are the type of factor. If not sure, apply `as.factor()` on those variables after reading the dataset.

```{r df1, eval=TRUE, echo=TRUE}
df1 <- read.csv(file.path("data", "5_BatteryLife.csv"))
df1$MaterialType <- as.factor(df1$MaterialType)
df1$Temperature <- as.factor(df1$Temperature)
```


Use boxplots to observe the differences of `LifeTime` among three levels of `MaterialType`, and, three levels of `Temperature`. 
We can observe that the average `LifeTime` tends to be lower for higher `Temperature`. 


```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=8}
# Draw the grouped boxplot
par(mfrow = c(1, 2))
boxplot(LifeTime ~ MaterialType, data = df1)
boxplot(LifeTime ~ Temperature, data = df1)
```
```{r, eval=TRUE, echo=FALSE}
par(mfrow = c(1, 1))
```

The effect model
\begin{equation}
y_{ijk} = \mu + \tau_i + \beta_j + (\tau\beta)_{ij} + \varepsilon_{ijk}
\end{equation}

- $\tau_i$ is the effect of the $i$th `MaterialType` level, $i = 1, 2, 3$.
- $\beta_j$ is the effect of the $j$th `Temperature` level, $j = 1, 2, 3$.
- $(\tau\beta)_{ij}$ is the interaction effect of the $i$th `MaterialType` level and the $j$th `Temperature` level.
- $\varepsilon_{ijk}$ is the random error, $k = 1, 2, 3, 4$, satisfying
$$\varepsilon_{ijk}\overset{i.i.d.}{\sim}N(0, \sigma^2) \text{ where } \sigma^2 \text{ is the conatnt variance.}$$

Three statistical hypotheses of this problem are defined as

$H_0$: There is no effect on the choice of `MaterialType`.

$H_0$: There is no `Temperature` effect.

$H_0$: There is no interaction effect between `MaterialType` and `Temperature`


The function `aov()` fits the ANOVA model, and the ANOVA table is obtained by calling `summary()`. 
One the left-hand-side of the R model formula `Y ~ X`, input the name of the response variable, i.e. `LifeTime`.
For factorial design, we test for the significance of the existence of the main effects as well as the the existence of the interaction effects.
In R model formula, the syntax of the interaction term is `X1:X2`.
In this battery life experiment, there are two factors, and hence the ANOVA model considers two main effects and one two-factor interaction.
On the right-hand-side of the R model formula, the following two inputs are the same:

- Separately input main effects and two-factor interaction, `MaterialType + Temperature + MaterialType:Temperature`, 
- Use multiplication `*` to include all interaction terms of the variables in the formula, `MaterialType * Temperature`.

```{r fit1, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit1 <- aov(LifeTime ~ MaterialType * Temperature, data = df1)
summary(fit1)
```

The p-values of both main effects and the interaction are less than the pre-specified significant level 0.05. That is, `MaterialType` and `Temperature` are both significantly related to the battery's `LifeTime`, and, the interaction of `MaterialType` and `Temperature` is also significant.

To visualize the analysis result of the factorial experiment, interaction plot is commonly used tool.
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
interaction.plot(
  x.factor = df1$Temperature, # x-axis variable
  trace.factor = df1$MaterialType, # variable for lines
  response = df1$LifeTime, # y-axis variable
  ylab = "LifeTime", xlab = "Temperature", 
  col = c("firebrick", "dodgerblue2", "forestgreen"),
  lty = 1, lwd = 3, trace.label = "MaterialType"
)
```
The plot shows two important conclusions:

- AT low temperature, $15^\circ C$, the lifetime of the battery is generally longer than those battery's in $125^\circ C$ environment. Among all materials, the life time of battery of type 2 material is the longest.
- AT middle temperature, $70^\circ C$, the lifetime of the battery of type 3 material is the longest.


The procedure of diagnosing the residual is similar to that for the one-way ANOVA model. Please refer to the handout of R codes in Chapter 3 for more details of interpreting the residual plots.
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=8, fig.height=8}
par(mfrow = c(2, 2))
plot(fit1)
# par(mfrow = c(1, 1))
```
```{r, eval=TRUE, echo = FALSE}
par(mfrow = c(1, 1))
```



One additional plot is to draw the scatter plot of the residual against the levels of each factor. A lack of any visually obvious pattern in the dots on the plot is desired.
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
plot(
  as.numeric(df1$MaterialType), fit1$residuals,
  xlab = "MaterialType", ylab = "residuals"
)
plot(
  as.numeric(df1$Temperature), fit1$residuals,
  xlab = "Temperature", ylab = "residuals"
)
```

Multiple comparison is performed for the treatment effect. The following codes demonstrate the use of Tukey’s test and Fisher's LSD method. 

For Tukey’s test, add the input argument `which = c("MaterialType", "Temperature")` to only show the test results of comparing differences among the `ExtPressure` levels. 

For Fisher's LSD method, specify `trt = c("MaterialType", "Temperature")` as the input argument to the `LSD.test()` function to show the comparison results among the `ExtPressure` levels. For information of interpreting the results, please refer to the handout of R codes in Chapter 3.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
TukeyHSD(fit1, which = c("MaterialType", "Temperature"))
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
if (!("agricolae" %in% rownames(installed.packages()))) {
  install.packages("agricolae")  
}
library(agricolae)
out <- LSD.test(fit1, trt = c("MaterialType", "Temperature"), p.adj = "bonferroni")
out$group
```

To fit the response surface model (RSM), the quantitative factor `Temperature` should be changed as of numeric type.
```{r df1q, eval=TRUE, echo=TRUE}
df1q <- read.csv(file.path("data", "5_BatteryLife.csv"))
df1q$MaterialType <- as.factor(df1q$MaterialType)
# Set MaterialType's dummy variable to use values -1, 0, 1
contrasts(df1q$MaterialType) <- contr.sum(3)
# Check the result of the model matrix of main effects
model.matrix( ~ MaterialType + Temperature, data = df1q)
```


The `lm()` function is used to fit the response surface model.

\begin{align*}
y & = \beta_0 + \beta_{1a} x_{1a} + \beta_{1b} x_{1b} + \beta_2 x_2 + \beta_{22} x_2^2  \\ 
  & + \beta_{1a2} x_{1a} x_2 + \beta_{1b2} x_{1b} x_2 + \beta_{1a22} x_{1a} x_2^2 + \beta_{1b22} x_{1b} x_2^2 \\
  & + \beta_{222} x_2^3 + \varepsilon
\end{align*}
where $x_{1\cdot}$ and $x_2$ are the value of `MaterialType` and `Temperature` respectively, and $\beta$'s are model coefficients.

In R model formula, the syntax indicating the higher order of the explanatory variable is `I(X^p)` where $p$ is the power. The RSM is
```{r ols1, eval=TRUE, echo=TRUE}
ols1 <- lm(
  LifeTime ~ (Temperature + I(Temperature^2)) * MaterialType + I(Temperature^3), 
  data = df1q
)
```

The `summary()` function for `lm` object is used to show the estimate of the coefficients and their significances. The coefficient estimate of $I(Temperature^3)$ is `NA` value because this cubic effect is aliased to the main effect.

```{r, eval=TRUE, echo=TRUE}
summary(ols1)
```

The Result of the RSM:

For Material Type 1
$$\text{Life} = + 169.380 - 2.489 \times \text{Temp} + 0.0129 \times \text{Temp}^2$$

For Material Type 2
$$\text{Life} = + 159.624 - 0.179 \times \text{Temp} + 0.4163 \times \text{Temp}^2$$
For Material Type 3
$$\text{Life} = + 132.762 + 0.893 \times \text{Temp} - 0.4322 \times \text{Temp}^2$$

## Example 5.2 Tool Life Experiment (Two Quantitative Factors)

Read the csv file `5_ToolLife.csv` in R. Make variables `TotalAngle` and `CuttingSpeed` to be the type of factor.

```{r df2, eval=TRUE, echo=TRUE}
df2 <- read.csv(file.path("data", "5_ToolLife.csv"))
df2$TotalAngle <- as.factor(df2$TotalAngle)
df2$CuttingSpeed <- as.factor(df2$CuttingSpeed)
```

The effect model
\begin{equation}
y_{ijk} = \mu + \tau_i + \beta_j + (\tau\beta)_{ij} + \varepsilon_{ijk}
\end{equation}

- $\tau_i$ is the effect of the $i$th `TotalAngle` level, $i = 1, 2, 3$.
- $\beta_j$ is the effect of the $j$th `CuttingSpeed` level, $j = 1, 2, 3$.
- $(\tau\beta)_{ij}$ is the interaction effect of the $i$th `TotalAngle` level and the $j$th `CuttingSpeed` level.
- $\varepsilon_{ijk}$ is the random error, $k = 1, 2$, satisfying
$$\varepsilon_{ijk}\overset{i.i.d.}{\sim}N(0, \sigma^2) \text{ where } \sigma^2 \text{ is the conatnt variance.}$$

Three statistical hypotheses of this problem are defined as

$H_0$: There is no `TotalAngle` effect.

$H_0$: There is no `CuttingSpeed` effect.

$H_0$: There is no interaction effect between `TotalAngle` and `CuttingSpeed`


Fit the ANOVA model by `aov()` function, and then print the ANOVA table by calling `summary()`. 

```{r, eval=TRUE, echo=TRUE}
fit2 <- aov(ToolLife ~ TotalAngle * CuttingSpeed, data = df2)
summary(fit2)
```

The p-values of both main effects and the interaction are less than the pre-specified significant level 0.05. That is, `TotalAngle` and `CuttingSpeed` are both significantly related to the battery's `ToolLife`, and, the interaction of `TotalAngle` and `CuttingSpeed` is also significant.

Hereafter, the residual checking and multiple comparison processes are left for practice.

The `lm()` function is used to fit the response model.

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1 x_2 + \beta_{11} x_1^2 + \beta_{22} x_2^2 + 
\varepsilon$$
where $x_1$ and $x_2$ are the value of `TotalAngle` and `CuttingSpeed` respectively, and $\beta$'s are model coefficients.

```{r ols2, eval=TRUE, echo=TRUE}
df2q <- read.csv(file.path("data", "5_ToolLife.csv"))
x1m <- mean(df2q$TotalAngle)
x2m <- mean(df2q$CuttingSpeed)
# Centralize the variables
df2q$TotalAngle <- df2q$TotalAngle - x1m
df2q$CuttingSpeed <- df2q$CuttingSpeed - x2m 
ols2 <- lm(
  ToolLife ~ TotalAngle*CuttingSpeed + I(TotalAngle^2) + I(CuttingSpeed^2), 
  data = df2q
)
summary(ols2)
```

Another choice of the response surface model for two factors is to include all possible interactions of all the second-order terms

\begin{align*}
y & = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1 x_2 + \beta_{11} x_1^2 + \beta_{22} x_2^2 \\
  & + \beta_{112} x_1^2 x_2 + \beta_{122} x_1 x_2^2 + \beta_{1122} x_1^2 x_2^2 + \varepsilon
\end{align*}

```{r, eval=TRUE, echo=TRUE}
ols2_full <- lm(
  ToolLife ~ (TotalAngle + I(TotalAngle^2))*(CuttingSpeed + I(CuttingSpeed^2)), 
  data = df2q
)
summary(ols2_full)
```

Draw the contour plot of the RSM. From the plot, we can conclude that setting mid-level of cutting speed and large angle could achieve higher tool life.
```{r ols2_full_contour, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
x1_grid <- seq(min(df2q$TotalAngle), max(df2q$TotalAngle), length = 100)
x2_grid <- seq(min(df2q$CuttingSpeed), max(df2q$CuttingSpeed), length = 100)
newx <- data.frame(
  TotalAngle = rep(x1_grid, each = 100),
  CuttingSpeed = rep(x2_grid, time = 100)
)
rs <- predict(ols2_full, newx)

rsplot_data <- data.frame(newx, rs = rs)
rsplot_data$TotalAngle <- rsplot_data$TotalAngle + x1m
rsplot_data$CuttingSpeed <- rsplot_data$CuttingSpeed + x2m

library(ggplot2)
#- Add color to the contour plot
ggplot(rsplot_data) +
  geom_contour(aes(TotalAngle, CuttingSpeed, z = rs), colour = "white") +
  geom_contour_filled(aes(TotalAngle, CuttingSpeed, z = rs))  

```

As a complementation, here are the codes for 3D plot of the RSM
```{r ols2_full_plotly, eval=FALSE, echo=TRUE, fig.align='center', fig.width=8, fig.height=8}
library(plotly)
library(htmlwidgets)
rsplot_matrix <- matrix(rsplot_data$rs, 100, 100)
p <- plot_ly(z = rsplot_matrix, type = "surface") %>%
  layout(scene = list(
      xaxis = list(
        title = 'TotalAngle',
        ticktext = lapply(seq(0, 100, 20), function(i) {
          diff(range(rsplot_data$TotalAngle))*i/100 + min(rsplot_data$TotalAngle)
        }), 
        tickvals = list(0, 20, 40, 60, 80, 100),
        tickmode = "array"
      ),
      yaxis = list(
        title = 'CuttingSpeed',
        ticktext = lapply(seq(0, 100, 20), function(i) {
          diff(range(rsplot_data$CuttingSpeed))*i/100 + min(rsplot_data$CuttingSpeed)
        }), 
        tickvals = list(0, 20, 40, 60, 80, 100),
        tickmode = "array"        
      ),
      zaxis = list(title = 'hat(ToolLife)')))

htmlwidgets::saveWidget(as_widget(p), "plotly_rsm_ch5.html")

```


## One Observation per Cell

Read the csv file `5_Impurity.csv` in R. Make variables `Temperature` and `Pressure` to be the type of factor.

```{r, eval=TRUE, echo=TRUE}
df3 <- read.csv(file.path("data", "5_Impurity.csv"))
df3$Temperature <- as.factor(df3$Temperature)
df3$Pressure <- as.factor(df3$Pressure)
```

The effect model
\begin{equation}
y_{ijk} = \mu + \tau_i + \beta_j + (\tau\beta)_{ij} + \varepsilon_{ijk}
\end{equation}

- $\tau_i$ is the effect of the $i$th `Temperature` level, $i = 1, 2, 3$.
- $\beta_j$ is the effect of the $j$th `Pressure` level, $j = 1, 2, 3$.
- $(\tau\beta)_{ij}$ is the interaction effect of the $i$th `Temperature` level and the $j$th `Pressure` level.
- $\varepsilon_{ijk}$ is the random error, $k = 1, 2$, satisfying
$$\varepsilon_{ijk}\overset{i.i.d.}{\sim}N(0, \sigma^2) \text{ where } \sigma^2 \text{ is the conatnt variance.}$$

Three statistical hypotheses of this problem are defined as

$H_0$: There is no `Temperature` effect.

$H_0$: There is no `Pressure` effect.

$H_0$: There is no interaction effect between `Temperature` and `Pressure`


Fit the ANOVA model by `aov()` function, and then print the ANOVA table by calling `summary()`. 
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit3 <- aov(Impurity ~ Temperature * Pressure, data = df3)
summary(fit3)
```

**(Important)** Because there is no replicates for each treatment combination, the ANOVA table does not exist given that the error variance $\sigma^2$ cannot be estimated.

Thus, for no-replicate scenario, we can only test for the two main effects.

$H_0$: There is no `Temperature` effect.

$H_0$: There is no `Pressure` effect.


```{r fit3_m, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit3_m <- aov(Impurity ~ Temperature + Pressure, data = df3)
summary(fit3_m)
```

Then the significance of the interaction effect is verified using Tukey's test of nonadditivity.
To implement Tukey's test of nonadditivity, the function `nonadditivity()` in the **agricolae** package is used. The resulting ANOVA for nonadditivity is shown below.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
library(agricolae)
naddtest <- nonadditivity(
  df3$Impurity, df3$Temperature, df3$Pressure, 
  df = df.residual(fit3_m), MSerror = deviance(fit3_m)/df.residual(fit3_m)
)
naddtest$ANOVA
```

The p-value of the nonadditivity is larger than the significance level 0.05 suggesting that there is no two-factor interaction of `Temperature` and `Pressure`.
05


## Three-Factor Factorial Experiment


Read the csv file `5_Impurity.csv` in R. Make variables `Carbonation`, `Pressure` and `LineSpeed` to be the type of factor.

```{r, eval=TRUE, echo=TRUE}
df4 <- read.csv(file.path("data", "5_SoftDrinkBottling.csv"))
df4$Carbonation <- as.factor(df4$Carbonation)
df4$Pressure <- as.factor(df4$Pressure)
df4$LineSpeed <- as.factor(df4$LineSpeed)
```

The effect model
\begin{equation}
y_{ijkl} = \mu + \tau_i + \beta_j + \gamma_k + (\tau\beta)_{ij} + (\tau\gamma)_{ik} + (\beta\gamma)_{jk} + (\tau\beta\gamma)_{ijk} + \varepsilon_{ijkl}
\end{equation}

- $\tau_i$ is the effect of the $i$th `Carbonation` level, $i = 1, 2, 3$.
- $\beta_j$ is the effect of the $j$th `Pressure` level, $j = 1, 2$.
- $\gamma_k$ is the effect of the $k$th `LineSpeed` level, $j = 1, 2$.
- $(\tau\beta)_{ij}$ are two-factor interactions.
- $(\tau\beta)_{ij}$, $(\tau\gamma)_{ik}$ and $(\beta\gamma)_{jk}$ are two-factor interactions.
- $(\tau\beta\gamma)_{ijk}$ is the three-factor interatoion.
- $\varepsilon_{ijkl}$ is the random error, $k = 1, 2$, satisfying
$$\varepsilon_{ijkl}\overset{i.i.d.}{\sim}N(0, \sigma^2) \text{ where } \sigma^2 \text{ is the conatnt variance.}$$

Totally, there are 7 statistical hypotheses of this problems
Fit the ANOVA model by `aov()` function, and then print the ANOVA table by calling `summary()`. 

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit4 <- aov(FillHeightsDev ~ Carbonation * Pressure * LineSpeed, data = df4)
summary(fit4)
```
The interpretation of these results are left for practice.

We can further remove all the terms with large p-value and then fit a reduced ANOVA model.
\begin{equation}
y_{ijkl} = \mu + \tau_i + \beta_j + \gamma_k + (\tau\beta)_{ij} + \varepsilon_{ijkl}
\end{equation}

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit4_r <- aov(
  FillHeightsDev ~ Carbonation + Pressure + LineSpeed + Carbonation:Pressure, 
  data = df4
)
summary(fit4_r)
```


\newpage
# Chapter 6: Two-Level Factorial Designs

## $2^2$-Design: Chemical Process Example
 
Read the csv file `6_ChemicalRecovery.csv` in R.
```{r, eval=TRUE, echo=TRUE}
df5 <- read.csv(file.path("data", "6_ChemicalRecovery.csv"))
```


The estimation of factor effects are:
```{r, eval=FALSE, echo=TRUE, fig.align='center', fig.width=5}
# Effect of factor A (reactant concentration)
mean(df5$Recovery[df5$ReactConc == 1] - df5$Recovery[df5$ReactConc == -1])
# or 2*mean(df5$Recovery*df5$ReactConc)

# Effect of factor A (catalyst amount)
mean(df5$Recovery[df5$CataAmo == 1] - df5$Recovery[df5$CataAmo == -1])

# Effect of two-factor interaction AB
twofi <- as.numeric(df5$ReactConc)*df5$CataAmo
mean(df5$Recovery[twofi == 1] - df5$Recovery[twofi == -1])
```
```{r, eval=TRUE, echo=FALSE, fig.align='center', fig.width=5}
# Effect of factor A (reactant concentration)
cat(sprintf("Effect of factor A (reactant concentration): %.4f\n\n", 
  mean(df5$Recovery[df5$ReactConc == 1] - df5$Recovery[df5$ReactConc == -1])
))
# Effect of factor A (catalyst amount)
cat(sprintf("Effect of factor A (catalyst amount): %.4f\n\n", 
  mean(df5$Recovery[df5$CataAmo == 1] - df5$Recovery[df5$CataAmo == -1])
))
# Effect of two-factor interaction AB
twofi <- as.numeric(df5$ReactConc)*df5$CataAmo
cat(sprintf("Effect of two-factor interaction AB: %.4f\n\n", 
  mean(df5$Recovery[twofi == 1] - df5$Recovery[twofi == -1])
))
```


Use `aov()` to fit the ANOVA model (model description in math is omitted).
Here, if the columns of the factor in the `data.frame` are not of the `factor` type. We can also specify the type of the factors as `factor` in the R model formula.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit5 <- aov(Recovery ~ factor(ReactConc)*factor(CataAmo), data = df5)
summary(fit5)
```

The two main effects are significant and their two-factor interaction may not exist.

To visualize the analysis result of the factorial experiment, interaction plot is commonly used tool.
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
interaction.plot(
  x.factor = df5$ReactConc, # x-axis variable
  trace.factor = df5$CataAmo, # variable for lines
  response = df5$Recovery, # y-axis variable
  ylab = "Recovery", xlab = "Reactant Concentration", 
  col = c("firebrick", "dodgerblue2"),
  lty = 1, lwd = 3, trace.label = "Catalyst\nAmount"
)
```


## $2^3$-Design: Plasma Etching Example

Read the csv file `6_PlasmaEtching_2^3.csv` in R.
```{r, eval=TRUE, echo=TRUE}
df6 <- read.csv(file.path("data", "6_PlasmaEtching_2^3.csv"))
```

The estimation of factor effects are:
```{r, eval=FALSE, echo=TRUE, fig.align='center', fig.width=5}
# Effect of main effects
2*mean(df6$EachRate*df6$Gap)
2*mean(df6$EachRate*df6$GasFlow)
2*mean(df6$EachRate*df6$Power)

# Effect of two-factor interactions
GapGasFlow <- df6$Gap * df6$GasFlow
GapPower <- df6$Gap * df6$Power
GasFlowPower <- df6$GasFlow * df6$Power
2*mean(df6$EachRate*GapGasFlow)
2*mean(df6$EachRate*GapPower)
2*mean(df6$EachRate*GasFlowPower)

# Effect of three-factor interaction
GapGasFlowPower <- df6$Gap * df6$GasFlow * df6$Power
2*mean(df6$EachRate*GapGasFlowPower)
```

```{r df6eff, eval=TRUE, echo=FALSE, fig.align='center', fig.width=5}
# Effect of two-factor interactions
GapGasFlow <- df6$Gap * df6$GasFlow
GapPower <- df6$Gap * df6$Power
GasFlowPower <- df6$GasFlow * df6$Power
# Effect of two-factor interaction AB
GapGasFlowPower <- df6$Gap * df6$GasFlow * df6$Power

effs <- c(
  2*mean(df6$EachRate*df6$Gap),
  2*mean(df6$EachRate*df6$GasFlow),
  2*mean(df6$EachRate*df6$Power),
  2*mean(df6$EachRate*GapGasFlow),
  2*mean(df6$EachRate*GapPower),
  2*mean(df6$EachRate*GasFlowPower),
  2*mean(df6$EachRate*GapGasFlowPower)
)

eff6 <- data.frame(
  Factor = c("A", "B", "C", "AB", "AC", "BC", "ABC"),
  "Est.Effect" = effs
)
print(eff6)
```

Use `aov()` to fit the ANOVA model (model description in math is omitted).
Here, if the columns of the factor in the `data.frame` are not of the `factor` type. We can also specify the type of the factors as `factor` in the R model formula.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit6 <- aov(
  EachRate ~ factor(Gap) * factor(GasFlow) * factor(Power), 
  data = df6
)
summary(fit6)
```

The significant effects are main effects `Gap` and `Power`, and, the two-factor interaction `Gap:Power`.

WE can also fit a response surface model and obtain the same conslusion.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
ols6_full <- lm(EachRate ~ Gap * GasFlow * Power, data = df6)
summary(ols6_full)
```


The final RSM is the reduced model that those insignificant terms are removed from the full model. Now we can use the real value of the factot levels as the predictors' values fro this reduced model for better interpretation.
```{r ols6_reduced, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
ols6_reduced <- lm(EachRate ~ NumGap * NumPower, data = df6)
summary(ols6_reduced)
```

```{r ols6_reduced_contour, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
x1_grid <- seq(min(df6$NumGap), max(df6$NumGap), length = 100)
x2_grid <- seq(min(df6$NumPower), max(df6$NumPower), length = 100)
newx <- data.frame(
  NumGap = rep(x1_grid, each = 100),
  NumPower = rep(x2_grid, time = 100)
)

rsplot_data6 <- data.frame(newx, rs = predict(ols6_reduced, newx))

library(ggplot2)
#- Add color to the contour plot
ggplot(rsplot_data6) +
  geom_contour(aes(NumGap, NumPower, z = rs), colour = "white") +
  geom_contour_filled(aes(NumGap, NumPower, z = rs))  
```

## Unreplicated $2^k$ Factorial Designs: The Resin Plant Experiment

Read the csv file `6_PilotPlant.csv` in R.
```{r, eval=TRUE, echo=TRUE}
df7 <- read.csv(file.path("data", "6_PilotPlant.csv"))
```

The estimation of factor effects are:
```{r eff7, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
# Compute the model matrix of all effect terms without intercept
mmat7 <- model.matrix( ~ Temperature*Pressure*CH2OConc*StirRate - 1, data = df7)
# Calculate the effect sizes using the +/- signs of the model matrix
eff7 <- numeric(ncol(mmat7))
for (i in 1:ncol(mmat7)) {
  eff7[i] <- 2*mean(df7$FiltrationRate*mmat7[,i])
}
names(eff7) <- colnames(mmat7)
```

```{r, eval=TRUE, echo=FALSE, fig.align='center', fig.width=5}
print(data.frame(
  Factor = names(eff7),
  "Est.Effect" = as.vector(eff7)
))
```

Because there is no replicates in each treatment combination, the estimate of the random error $\sigma^2$ does not exist and the ANOVA table is not available. Of all effect terms, we try of eliminate some of them before analyzing the data. According to the model assumption, the effects that are negligible should be similar to random error which is normally distributed with zero mean and constant variance. Therefore, a QQ-plot or half-Normal plot is helpful to identify effective effects. Belows are the codes of half-Normal plot.

```{r halfqqnorm, eval=TRUE, echo=TRUE}
# Half Normal Plot
halfqqnorm <- function(input, tol = 0.5) {
	y <- sort(abs(input))
	nq <- qnorm(seq(0.5, 0.99, length = length(y)))
	plot(y, nq, yaxt = "n", pch = 15, 
	     xlab = "|Effects|", ylab = "Half Normal Probability")
	title("Half Normal Plot")
	# choose anchor point to draw a straight line
	s <- min(which(diff(y)/diff(range(y)) > 1/(length(y)-1)))
	abline(a = -y[1]*(nq[s]-nq[1])/(y[s]-y[1]), b = (nq[s]-nq[1])/(y[s]-y[1]))
	axis(2, at = qnorm(seq(0.5, 0.9999, length = 5)), 
	     labels = round(seq(0, 1, length = 5),2))
	loc <- sqrt((nq - (y - y[1])*(nq[s]-nq[1])/(y[s]-y[1]))^2) > tol
	if (is.null(names(y))) {
		text(y[loc], nq[loc], order(abs(input))[loc], pos = 2)
	} else {
		text(y[loc], nq[loc], names(abs(input))[order(abs(input))[loc]], pos = 2)
	}
}
```


```{r h7, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
halfqqnorm(eff7)
```

By the half Normal plot, we find out that the main effects `Temperature`, `CH2OConc`and `StirRate` and interactions `Temperature:CH2OConc`, `Temperature:StirRate` appear to be large.

Based on the observation from the half Normal plot, now the ANOVA model is

\begin{equation}
y_{ijkl} = \mu + \tau_i + \beta_j + \gamma_k + (\tau\beta)_{ij} + (\tau\gamma)_{ik} + \varepsilon_{ijkl}
\end{equation}

- $\tau_i$ is the effect of the $i$th `Temperature` level, $i = 1, 2$.
- $\beta_j$ is the effect of the $j$th `CH2OConc` level, $j = 1, 2$.
- $\gamma_k$ is the effect of the $k$th `StirRate` level, $j = 1, 2$.
- $(\tau\beta)_{ij}$ is the interaction effect of the $i$th `Temperature` level and the $j$th `CH2OConc` level.
- $(\tau\gamma)_{ij}$ is the interaction effect of the $i$th `Temperature` level and the $k$th `StirRate` level.
- $\varepsilon_{ijkl}$ is the random error, $l = 1, 2$, satisfying
$$\varepsilon_{ijkl}\overset{i.i.d.}{\sim}N(0, \sigma^2) \text{ where } \sigma^2 \text{ is the conatnt variance.}$$

Use `aov()` to fit the ANOVA model.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit7 <- aov(
  FiltrationRate ~ factor(Temperature) * (factor(CH2OConc) + factor(StirRate)), 
  data = df7
)
summary(fit7)
```

Here, we left the interpretation of the response surface model for practice.

<!--
```{r, eval=TRUE, echo=TRUE}
df8 <- read.csv(file.path("data", "6_Drilling.csv"))
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit8 <- aov(AdvRate ~ factor(FlowRate) * (factor(RotatSp) + factor(MudType)), data = df8)
summary(fit8)
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
interaction.plot(
  x.factor = df8$FlowRate, # x-axis variable
  trace.factor = df8$RotatSp, # variable for lines
  response = df8$AdvRate, # y-axis variable
  ylab = "AdvRate", xlab = "FlowRate", 
  col = c("firebrick", "dodgerblue2"),
  lty = 1, lwd = 3, trace.label = "RotatSp"
)

interaction.plot(
  x.factor = df8$FlowRate, # x-axis variable
  trace.factor = df8$MudType, # variable for lines
  response = df8$AdvRate, # y-axis variable
  ylab = "AdvRate", xlab = "FlowRate", 
  col = c("firebrick", "dodgerblue2"),
  lty = 1, lwd = 3, trace.label = "MudType"
)
```

```{r, eval=TRUE, echo=TRUE}
df9 <- read.csv(file.path("data", "6_Sidewall.csv"))
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit9 <- aov(AvgDefectNum ~ factor(Temperature) + factor(ResinFlow), data = df9)
summary(fit9)
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=6, fig.height=6}
par(mfrow = c(2, 2))
plot(fit9)
par(mfrow = c(1, 1))
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
plot(
  as.numeric(df9$ClampTime), fit9$residuals, pch = 16,
  xlab = "ClampTime", ylab = "residuals", xlim = c(-2, 2)
)
abline(h = 0, lty = 3)
```

```{r, eval=TRUE, echo=TRUE}
df10 <- read.csv(file.path("data", "6_Wafer.csv"))
```

```{r, eval=TRUE, echo=TRUE}
df11 <- read.csv(file.path("data", "6_CreditCardMarketing.csv"))
```
-->

## Addition of Center Points to a $2^k$ Designs 

Recall the Resin Plant Experiment. Read the csv file `6_PilotPlant.csv` in R.

```{r df7a, eval=TRUE, echo=TRUE}
df7 <- read.csv(file.path("data", "6_PilotPlant.csv"))
```

```{r df7c, eval=TRUE, echo=TRUE}
df7_C <- data.frame(
  Temperature = rep(0, 4),
  Pressure = rep(0, 4),
  CH2OConc = rep(0, 4),
  StirRate = rep(0, 4),
  FiltrationRate = c(73, 75, 66, 69)
)
```

The calculation of $SS_{Pure quadratic}$ with degree of freedom 1.
```{r sspq, eval=TRUE, echo=TRUE}
Yf_bar <- mean(df7$FiltrationRate)
Yc_bar <- mean(df7_C$FiltrationRate)
nf <- nrow(df7)
nc <- nrow(df7_C)
SS_pureQuad <- nf*nc*(Yf_bar - Yc_bar)^2/(nf + nc)
```

The calculation of $SS_{E}$ with degree of freedom $n_c - 1$.

```{r sse, eval=TRUE, echo=TRUE}
SS_E <- sum((df7_C$FiltrationRate - mean(df7_C$FiltrationRate))^2)
```

To test the significance of the Curvature, we compute the ratio of $MS_{Pure quadratic}$ and $MS_{E}$ as the $F$-statistic which follows the $F$ distribution with degrees of freedom 1 and $n_c - 1$.
```{r, eval=TRUE, echo=TRUE}
MS_pureQuad <- SS_pureQuad/1
MS_E <- SS_E/(nc - 1)
fval <- MS_pureQuad/MS_E
pval <- 1 - pf(fval, 1, nc - 1) # p-value
```

The part of the testing the significance of the Curvature of the ANOVA table.
```{r, eval=TRUE, echo=TRUE}
print(data.frame(
  Source = c("Curvature", "Residual"),
  SS = c(SS_pureQuad, SS_E),
  DF = c(1, nc - 1), 
  MS = c(MS_pureQuad, MS_E),
  "F" = c(sprintf("%.3f", fval), ""),
  "Pr(>F)" = c(sprintf("%.3f", pval), "")
))
```


<!--

Demo in class
```{r, eval=TRUE, echo=TRUE}
df7 <- read.csv(file.path("data", "6_PilotPlant.csv"))
df7_C <- data.frame(
  Temperature = rep(0, 4),
  Pressure = rep(0, 4),
  CH2OConc = rep(0, 4),
  StirRate = rep(0, 4),
  FiltrationRate = c(73, 75, 66, 69)
)
```


```{r, eval=TRUE, echo=TRUE}
fit7 <- aov(FiltrationRate ~ factor(Temperature) * factor(Pressure) * factor(CH2OConc) * factor(StirRate), data = df7)
summary(fit7)
```

```{r, eval=TRUE, echo=TRUE}
df7A <- rbind(df7, df7_C)
fit7A <- aov(
  FiltrationRate ~ factor(Temperature)*factor(Pressure)*factor(CH2OConc)*factor(StirRate), 
  data = df7A
)
summary(fit7A)
```
-->