---
title: "Ch-05 R Codes"
author: "Ping-Yang Chen"
#date: "2024-03-22"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Textbook: Montgomery, D. C. (2012). *Design and analysis of experiments*, 8th Edition. John Wiley & Sons.

Online handouts: https://github.com/PingYangChen/ANOVA_Course_R_Code

```{r, eval=FALSE, echo=FALSE}
demo1 <- read.csv(file.path("data", "5_TwoFactorDesign.csv"))
demo2 <- read.csv(file.path("data", "5_TwoFactorDesignInteraction.csv"))
demo1$FactorA <- as.factor(demo1$FactorA)
demo1$FactorB <- as.factor(demo1$FactorB)
demo2$FactorA <- as.factor(demo2$FactorA)
demo2$FactorB <- as.factor(demo2$FactorB)
```

## Example 5.1 The Battery Life Experiment

Read the csv file `5_BatteryLife.csv` in R. Make sure that in the `data.frame` the variables `MaterialType` and `Temperature` are the type of factor. If not sure, apply `as.factor()` on those variables after reading the dataset.

```{r df1, eval=TRUE, echo=TRUE}
df1 <- read.csv(file.path("data", "5_BatteryLife.csv"))
df1$MaterialType <- as.factor(df1$MaterialType)
df1$Temperature <- as.factor(df1$Temperature)
```


Use boxplots to observe the differences of `LifeTime` among three levels of `MaterialType`, and, three levels of `Temperature`. 
We can observe that the average `LifeTime` tends to be lower for higher `Temperature`. 


```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=8}
# Draw the grouped boxplot
par(mfrow = c(1, 2))
boxplot(LifeTime ~ MaterialType, data = df1)
boxplot(LifeTime ~ Temperature, data = df1)
```
```{r, eval=TRUE, echo=FALSE}
par(mfrow = c(1, 1))
```

The effect model
\begin{equation}
y_{ijk} = \mu + \tau_i + \beta_j + (\tau\beta)_{ij} + \varepsilon_{ijk}
\end{equation}

- $\tau_i$ is the effect of the $i$th `MaterialType` level, $i = 1, 2, 3$.
- $\beta_j$ is the effect of the $j$th `Temperature` level, $j = 1, 2, 3$.
- $(\tau\beta)_{ij}$ is the interaction effect of the $i$th `MaterialType` level and the $j$th `Temperature` level.
- $\varepsilon_{ijk}$ is the random error, $k = 1, 2, 3, 4$, satisfying
$$\varepsilon_{ijk}\overset{i.i.d.}{\sim}N(0, \sigma^2) \text{ where } \sigma^2 \text{ is the conatnt variance.}$$

Three statistical hypotheses of this problem are defined as

$H_0$: There is no effect on the choice of `MaterialType`.

$H_0$: There is no `Temperature` effect.

$H_0$: There is no interaction effect between `MaterialType` and `Temperature`


The function `aov()` fits the ANOVA model, and the ANOVA table is obtained by calling `summary()`. 
One the left-hand-side of the R model formula `Y ~ X`, input the name of the response variable, i.e. `LifeTime`.
For factorial design, we test for the significance of the existence of the main effects as well as the the existence of the interaction effects.
In R model formula, the syntax of the interaction term is `X1:X2`.
In this battery life experiment, there are two factors, and hence the ANOVA model considers two main effects and one two-factor interaction.
On the right-hand-side of the R model formula, the following two inputs are the same:

- Separately input main effects and two-factor interaction, `MaterialType + Temperature + MaterialType:Temperature`, 
- Use multiplication `*` to include all interaction terms of the variables in the formula, `MaterialType * Temperature`.

```{r fit1, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit1 <- aov(LifeTime ~ MaterialType * Temperature, data = df1)
summary(fit1)
```

The p-values of both main effects and the interaction are less than the pre-specified significant level 0.05. That is, `MaterialType` and `Temperature` are both significantly related to the battery's `LifeTime`, and, the interaction of `MaterialType` and `Temperature` is also significant.

To visualize the analysis result of the factorial experiment, interaction plot is commonly used tool.
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
interaction.plot(
  x.factor = df1$Temperature, # x-axis variable
  trace.factor = df1$MaterialType, # variable for lines
  response = df1$LifeTime, # y-axis variable
  ylab = "LifeTime", xlab = "Temperature", 
  col = c("firebrick", "dodgerblue2", "forestgreen"),
  lty = 1, lwd = 3, trace.label = "MaterialType"
)
```
The plot shows two important conclusions:

- AT low temperature, $15^\circ C$, the lifetime of the battery is generally longer than those battery's in $125^\circ C$ environment. Among all materials, the life time of battery of type 2 material is the longest.
- AT middle temperature, $70^\circ C$, the lifetime of the battery of type 3 material is the longest.


The procedure of diagnosing the residual is similar to that for the one-way ANOVA model. Please refer to the handout of R codes in Chapter 3 for more details of interpreting the residual plots.
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=8, fig.height=8}
par(mfrow = c(2, 2))
plot(fit1)
# par(mfrow = c(1, 1))
```
```{r, eval=TRUE, echo = FALSE}
par(mfrow = c(1, 1))
```



One additional plot is to draw the scatter plot of the residual against the levels of each factor. A lack of any visually obvious pattern in the dots on the plot is desired.
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
plot(
  as.numeric(df1$MaterialType), fit1$residuals,
  xlab = "MaterialType", ylab = "residuals"
)
plot(
  as.numeric(df1$Temperature), fit1$residuals,
  xlab = "Temperature", ylab = "residuals"
)
```

Multiple comparison is performed for the treatment effect. The following codes demonstrate the use of Tukey’s test and Fisher's LSD method. 

For Tukey’s test, add the input argument `which = c("MaterialType", "Temperature")` to only show the test results of comparing differences among the `ExtPressure` levels. 

For Fisher's LSD method, specify `trt = c("MaterialType", "Temperature")` as the input argument to the `LSD.test()` function to show the comparison results among the `ExtPressure` levels. For information of interpreting the results, please refer to the handout of R codes in Chapter 3.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
TukeyHSD(fit1, which = c("MaterialType", "Temperature"))
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
if (!("agricolae" %in% rownames(installed.packages()))) {
  install.packages("agricolae")  
}
library(agricolae)
out <- LSD.test(fit1, trt = c("MaterialType", "Temperature"), p.adj = "bonferroni")
out$group
```

To fit the response surface model (RSM), the quantitative factor `Temperature` should be changed as of numeric type.
```{r df1q, eval=TRUE, echo=TRUE}
df1q <- read.csv(file.path("data", "5_BatteryLife.csv"))
df1q$MaterialType <- as.factor(df1q$MaterialType)
# Set MaterialType's dummy variable to use values -1, 0, 1
contrasts(df1q$MaterialType) <- contr.sum(3)
# Check the result of the model matrix of main effects
model.matrix( ~ MaterialType + Temperature, data = df1q)
```


The `lm()` function is used to fit the response surface model.

\begin{align*}
y & = \beta_0 + \beta_{1a} x_{1a} + \beta_{1b} x_{1b} + \beta_2 x_2 + \beta_{22} x_2^2  \\ 
  & + \beta_{1a2} x_{1a} x_2 + \beta_{1b2} x_{1b} x_2 + \beta_{1a22} x_{1a} x_2^2 + \beta_{1b22} x_{1b} x_2^2 \\
  & + \beta_{222} x_2^3 + \varepsilon
\end{align*}
where $x_{1\cdot}$ and $x_2$ are the value of `MaterialType` and `Temperature` respectively, and $\beta$'s are model coefficients.

In R model formula, the syntax indicating the higher order of the explanatory variable is `I(X^p)` where $p$ is the power. The RSM is
```{r ols1, eval=TRUE, echo=TRUE}
ols1 <- lm(
  LifeTime ~ (Temperature + I(Temperature^2)) * MaterialType + I(Temperature^3), 
  data = df1q
)
```

The `summary()` function for `lm` object is used to show the estimate of the coefficients and their significances. The coefficient estimate of $I(Temperature^3)$ is `NA` value because this cubic effect is aliased to the main effect.

```{r, eval=TRUE, echo=TRUE}
summary(ols1)
```

The Result of the RSM:

For Material Type 1
$$\text{Life} = + 169.380 - 2.489 \times \text{Temp} + 0.0129 \times \text{Temp}^2$$

For Material Type 2
$$\text{Life} = + 159.624 - 0.179 \times \text{Temp} + 0.4163 \times \text{Temp}^2$$
For Material Type 3
$$\text{Life} = + 132.762 + 0.893 \times \text{Temp} - 0.4322 \times \text{Temp}^2$$

## Example 5.2 Tool Life Experiment (Two Quantitative Factors)

Read the csv file `5_ToolLife.csv` in R. Make variables `TotalAngle` and `CuttingSpeed` to be the type of factor.

```{r df2, eval=TRUE, echo=TRUE}
df2 <- read.csv(file.path("data", "5_ToolLife.csv"))
df2$TotalAngle <- as.factor(df2$TotalAngle)
df2$CuttingSpeed <- as.factor(df2$CuttingSpeed)
```

The effect model
\begin{equation}
y_{ijk} = \mu + \tau_i + \beta_j + (\tau\beta)_{ij} + \varepsilon_{ijk}
\end{equation}

- $\tau_i$ is the effect of the $i$th `TotalAngle` level, $i = 1, 2, 3$.
- $\beta_j$ is the effect of the $j$th `CuttingSpeed` level, $j = 1, 2, 3$.
- $(\tau\beta)_{ij}$ is the interaction effect of the $i$th `TotalAngle` level and the $j$th `CuttingSpeed` level.
- $\varepsilon_{ijk}$ is the random error, $k = 1, 2$, satisfying
$$\varepsilon_{ijk}\overset{i.i.d.}{\sim}N(0, \sigma^2) \text{ where } \sigma^2 \text{ is the conatnt variance.}$$

Three statistical hypotheses of this problem are defined as

$H_0$: There is no `TotalAngle` effect.

$H_0$: There is no `CuttingSpeed` effect.

$H_0$: There is no interaction effect between `TotalAngle` and `CuttingSpeed`


Fit the ANOVA model by `aov()` function, and then print the ANOVA table by calling `summary()`. 

```{r, eval=TRUE, echo=TRUE}
fit2 <- aov(ToolLife ~ TotalAngle * CuttingSpeed, data = df2)
summary(fit2)
```

The p-values of both main effects and the interaction are less than the pre-specified significant level 0.05. That is, `TotalAngle` and `CuttingSpeed` are both significantly related to the battery's `ToolLife`, and, the interaction of `TotalAngle` and `CuttingSpeed` is also significant.

Hereafter, the residual checking and multiple comparison processes are left for practice.

The `lm()` function is used to fit the response model.

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1 x_2 + \beta_{11} x_1^2 + \beta_{22} x_2^2 + 
\varepsilon$$
where $x_1$ and $x_2$ are the value of `TotalAngle` and `CuttingSpeed` respectively, and $\beta$'s are model coefficients.

```{r ols2, eval=TRUE, echo=TRUE}
df2q <- read.csv(file.path("data", "5_ToolLife.csv"))
x1m <- mean(df2q$TotalAngle)
x2m <- mean(df2q$CuttingSpeed)
# Centralize the variables
df2q$TotalAngle <- df2q$TotalAngle - x1m
df2q$CuttingSpeed <- df2q$CuttingSpeed - x2m 
ols2 <- lm(
  ToolLife ~ TotalAngle*CuttingSpeed + I(TotalAngle^2) + I(CuttingSpeed^2), 
  data = df2q
)
summary(ols2)
```

Another choice of the response surface model for two factors is to include all possible interactions of all the second-order terms

\begin{align*}
y & = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1 x_2 + \beta_{11} x_1^2 + \beta_{22} x_2^2 \\
  & + \beta_{112} x_1^2 x_2 + \beta_{122} x_1 x_2^2 + \beta_{1122} x_1^2 x_2^2 + \varepsilon
\end{align*}

```{r, eval=TRUE, echo=TRUE}
ols2_full <- lm(
  ToolLife ~ (TotalAngle + I(TotalAngle^2))*(CuttingSpeed + I(CuttingSpeed^2)), 
  data = df2q
)
summary(ols2_full)
```

Draw the contour plot of the RSM. From the plot, we can conclude that setting mid-level of cutting speed and large angle could achieve higher tool life.
```{r ols2_full_contour, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
x1_grid <- seq(min(df2q$TotalAngle), max(df2q$TotalAngle), length = 100)
x2_grid <- seq(min(df2q$CuttingSpeed), max(df2q$CuttingSpeed), length = 100)
newx <- data.frame(
  TotalAngle = rep(x1_grid, each = 100),
  CuttingSpeed = rep(x2_grid, time = 100)
)
rs <- predict(ols2_full, newx)

rsplot_data <- data.frame(newx, rs = rs)
rsplot_data$TotalAngle <- rsplot_data$TotalAngle + x1m
rsplot_data$CuttingSpeed <- rsplot_data$CuttingSpeed + x2m

library(ggplot2)
#- Add color to the contour plot
ggplot(rsplot_data) +
  geom_contour(aes(TotalAngle, CuttingSpeed, z = rs), colour = "white") +
  geom_contour_filled(aes(TotalAngle, CuttingSpeed, z = rs))  

```

As a complementation, here are the codes for 3D plot of the RSM
```{r ols2_full_plotly, eval=FALSE, echo=TRUE, fig.align='center', fig.width=8, fig.height=8}
library(plotly)
library(htmlwidgets)
rsplot_matrix <- matrix(rsplot_data$rs, 100, 100)
p <- plot_ly(z = rsplot_matrix, type = "surface") %>%
  layout(scene = list(
      xaxis = list(
        title = 'TotalAngle',
        ticktext = lapply(seq(0, 100, 20), function(i) {
          diff(range(rsplot_data$TotalAngle))*i/100 + min(rsplot_data$TotalAngle)
        }), 
        tickvals = list(0, 20, 40, 60, 80, 100),
        tickmode = "array"
      ),
      yaxis = list(
        title = 'CuttingSpeed',
        ticktext = lapply(seq(0, 100, 20), function(i) {
          diff(range(rsplot_data$CuttingSpeed))*i/100 + min(rsplot_data$CuttingSpeed)
        }), 
        tickvals = list(0, 20, 40, 60, 80, 100),
        tickmode = "array"        
      ),
      zaxis = list(title = 'hat(ToolLife)')))

htmlwidgets::saveWidget(as_widget(p), "plotly_rsm_ch5.html")

```


## One Observation per Cell

Read the csv file `5_Impurity.csv` in R. Make variables `Temperature` and `Pressure` to be the type of factor.

```{r, eval=TRUE, echo=TRUE}
df3 <- read.csv(file.path("data", "5_Impurity.csv"))
df3$Temperature <- as.factor(df3$Temperature)
df3$Pressure <- as.factor(df3$Pressure)
```

The effect model
\begin{equation}
y_{ijk} = \mu + \tau_i + \beta_j + (\tau\beta)_{ij} + \varepsilon_{ijk}
\end{equation}

- $\tau_i$ is the effect of the $i$th `Temperature` level, $i = 1, 2, 3$.
- $\beta_j$ is the effect of the $j$th `Pressure` level, $j = 1, 2, 3$.
- $(\tau\beta)_{ij}$ is the interaction effect of the $i$th `Temperature` level and the $j$th `Pressure` level.
- $\varepsilon_{ijk}$ is the random error, $k = 1, 2$, satisfying
$$\varepsilon_{ijk}\overset{i.i.d.}{\sim}N(0, \sigma^2) \text{ where } \sigma^2 \text{ is the conatnt variance.}$$

Three statistical hypotheses of this problem are defined as

$H_0$: There is no `Temperature` effect.

$H_0$: There is no `Pressure` effect.

$H_0$: There is no interaction effect between `Temperature` and `Pressure`


Fit the ANOVA model by `aov()` function, and then print the ANOVA table by calling `summary()`. 
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit3 <- aov(Impurity ~ Temperature * Pressure, data = df3)
summary(fit3)
```

**(Important)** Because there is no replicates for each treatment combination, the ANOVA table does not exist given that the error variance $\sigma^2$ cannot be estimated.

Thus, for no-replicate scenario, we can only test for the two main effects.

$H_0$: There is no `Temperature` effect.

$H_0$: There is no `Pressure` effect.


```{r fit3_m, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit3_m <- aov(Impurity ~ Temperature + Pressure, data = df3)
summary(fit3_m)
```

Then the significance of the interaction effect is verified using Tukey's test of nonadditivity.
To implement Tukey's test of nonadditivity, the function `nonadditivity()` in the **agricolae** package is used. The resulting ANOVA for nonadditivity is shown below.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
library(agricolae)
naddtest <- nonadditivity(
  df3$Impurity, df3$Temperature, df3$Pressure, 
  df = df.residual(fit3_m), MSerror = deviance(fit3_m)/df.residual(fit3_m)
)
naddtest$ANOVA
```

The p-value of the nonadditivity is larger than the significance level 0.05 suggesting that there is no two-factor interaction of `Temperature` and `Pressure`.
05


## Three-Factor Factorial Experiment


Read the csv file `5_Impurity.csv` in R. Make variables `Carbonation`, `Pressure` and `LineSpeed` to be the type of factor.

```{r, eval=TRUE, echo=TRUE}
df4 <- read.csv(file.path("data", "5_SoftDrinkBottling.csv"))
df4$Carbonation <- as.factor(df4$Carbonation)
df4$Pressure <- as.factor(df4$Pressure)
df4$LineSpeed <- as.factor(df4$LineSpeed)
```

The effect model
\begin{equation}
y_{ijkl} = \mu + \tau_i + \beta_j + \gamma_k + (\tau\beta)_{ij} + (\tau\gamma)_{ik} + (\beta\gamma)_{jk} + (\tau\beta\gamma)_{ijk} + \varepsilon_{ijkl}
\end{equation}

- $\tau_i$ is the effect of the $i$th `Carbonation` level, $i = 1, 2, 3$.
- $\beta_j$ is the effect of the $j$th `Pressure` level, $j = 1, 2$.
- $\gamma_k$ is the effect of the $k$th `LineSpeed` level, $j = 1, 2$.
- $(\tau\beta)_{ij}$ are two-factor interactions.
- $(\tau\beta)_{ij}$, $(\tau\gamma)_{ik}$ and $(\beta\gamma)_{jk}$ are two-factor interactions.
- $(\tau\beta\gamma)_{ijk}$ is the three-factor interatoion.
- $\varepsilon_{ijkl}$ is the random error, $k = 1, 2$, satisfying
$$\varepsilon_{ijkl}\overset{i.i.d.}{\sim}N(0, \sigma^2) \text{ where } \sigma^2 \text{ is the conatnt variance.}$$

Totally, there are 7 statistical hypotheses of this problems
Fit the ANOVA model by `aov()` function, and then print the ANOVA table by calling `summary()`. 

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit4 <- aov(FillHeightsDev ~ Carbonation * Pressure * LineSpeed, data = df4)
summary(fit4)
```
The interpretation of these results are left for practice.

We can further remove all the terms with large p-value and then fit a reduced ANOVA model.
\begin{equation}
y_{ijkl} = \mu + \tau_i + \beta_j + \gamma_k + (\tau\beta)_{ij} + \varepsilon_{ijkl}
\end{equation}

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit4_r <- aov(
  FillHeightsDev ~ Carbonation + Pressure + LineSpeed + Carbonation:Pressure, 
  data = df4
)
summary(fit4_r)
```
