---
title: "Ch-05 & 06 R Codes"
author: "Ping-Yang Chen"
date: "2024-04-xx"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Textbook: Montgomery, D. C. (2012). *Design and analysis of experiments*, 8th Edition. John Wiley & Sons.

Online handouts: https://github.com/PingYangChen/ANOVA_Course_R_Code

# Chapter 5


```{r, eval=TRUE, echo=TRUE}
demo1 <- read.csv(file.path("data", "5_TwoFactorDesign.csv"))
demo2 <- read.csv(file.path("data", "5_TwoFactorDesignInteraction.csv"))
demo1$FactorA <- as.factor(demo1$FactorA)
demo1$FactorB <- as.factor(demo1$FactorB)
demo2$FactorA <- as.factor(demo2$FactorA)
demo2$FactorB <- as.factor(demo2$FactorB)
```

## Factorial Designs




Read the csv file `5_BatteryLife.csv` in R. Make sure that in the `data.frame` the variables `ExtPressure` and `ResinBatchNo` are the type of factor. If not sure, apply `as.factor()` on those variables after reading the dataset.

```{r, eval=TRUE, echo=TRUE}
df1 <- read.csv(file.path("data", "5_BatteryLife.csv"))
df1$MaterialType <- as.factor(df1$MaterialType)
df1$Temperature <- as.factor(df1$Temperature)
```


Use boxplots to observe the differences of `LifeTime` among three levels of `MaterialType`, and, three levels of `Temperature`. 

The average `LifeTime` tends to be lower for higher `Temperature`. 


```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=8}
# Draw the grouped boxplot
par(mfrow = c(1, 2))
boxplot(LifeTime ~ MaterialType, data = df1)
boxplot(LifeTime ~ Temperature, data = df1)
```
```{r, eval=TRUE, echo=FALSE}
par(mfrow = c(1, 1))
```


The function `aov()` fits the ANOVA model, and the ANOVA table is obtained by calling `summary()`. For Randomized Complete Block Design (RCBD), the R model formula includes two items on its right-hand-side: `ExtPressure + ResinBatchNo`. It's important to note that the order of placing the factor and block effects is somewhat arbitrary for RCBD. However, this order matters if the design is incomplete, as some sums of squares are adjusted accordingly.
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
fit1 <- aov(LifeTime ~ MaterialType * Temperature, data = df1)
summary(fit1)
```


```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=6, fig.height=6}
interaction.plot(
  x.factor = df1$Temperature, # x-axis variable
  trace.factor = df1$MaterialType, # variable for lines
  response = df1$LifeTime, # y-axis variable
  ylab = "LifeTime", xlab = "Temperature", 
  col = c("firebrick", "dodgerblue2", "forestgreen"),
  lty = 1, lwd = 3, trace.label = "MaterialType"
)
```



The procedure of diagnosing the residual is similar to that for the one-way ANOVA model. Please refer to the handout of R codes in Chapter 3 for more details of interpreting the residual plots.
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=8, fig.height=8}
par(mfrow = c(2, 2))
plot(fit1)
```
```{r, eval=TRUE, echo = FALSE}
par(mfrow = c(1, 1))
```





One additional plot is to draw the scatter plot of the residual against the levels of the block. A lack of any visually obvious pattern in the dots on the plot is desired.
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4, fig.height=4}
plot(
  as.numeric(df1$MaterialType), fit1$residuals,
  xlab = "MaterialType", ylab = "residuals"
)
plot(
  as.numeric(df1$Temperature), fit1$residuals,
  xlab = "Temperature", ylab = "residuals"
)
```
Multiple comparison is performed for the treatment effect. The following codes demonstrate the use of Tukey’s test and Fisher's LSD method. For Tukey’s test, add the input argument `which = "ExtPressure"` to only show the test results of comparing differences among the `ExtPressure` levels. For Fisher's LSD method, specify `trt = "ExtPressure"` as the input argument to the `LSD.test()` function to show the comparison results among the `ExtPressure` levels. For information of interpreting the results, please refer to the handout of R codes in Chapter 3.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
TukeyHSD(fit1, which = c("MaterialType", "Temperature"))
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
if (!("agricolae" %in% rownames(installed.packages()))) {
  install.packages("agricolae")  
}
library(agricolae)
out <- LSD.test(fit1, trt = c("MaterialType", "Temperature"), p.adj = "bonferroni")
out$group
```


```{r df1q, eval=TRUE, echo=TRUE}
df1q <- df1
df1q$Temperature <- as.numeric(df1q$Temperature) - 2
contrasts(df1q$MaterialType) <- contr.sum(3)
ols1 <- lm(LifeTime ~ (Temperature + I(Temperature^2)) * MaterialType + I(Temperature^3), data = df1q)
anova(ols1)
summary(ols1)
```

```{r, eval=TRUE, echo=TRUE}
summary(ols1)
```



```{r, eval=TRUE, echo=TRUE}
df2 <- read.csv(file.path("data", "5_ToolLife.csv"))
df2$TotalAngle <- as.factor(df2$TotalAngle)
df2$CuttingSpeed <- as.factor(df2$CuttingSpeed)
```

```{r, eval=TRUE, echo=TRUE}
fit2 <- aov(ToolLife ~ TotalAngle*CuttingSpeed, data = df2)
summary(fit2)
```



```{r, eval=TRUE, echo=TRUE}
df2q <- read.csv(file.path("data", "5_ToolLife.csv"))
x1m <- mean(df2q$TotalAngle)
x2m <- mean(df2q$CuttingSpeed)
df2q$TotalAngle <- df2q$TotalAngle - x1m
df2q$CuttingSpeed <- df2q$CuttingSpeed - x2m 
#df2q$TotalAngle <- 2*(df2q$TotalAngle - min(df2q$TotalAngle))/diff(range(df2q$TotalAngle)) - 1
#df2q$CuttingSpeed <- 2*(df2q$CuttingSpeed - min(df2q$CuttingSpeed))/diff(range(df2q$CuttingSpeed)) - 1
ols2 <- lm(ToolLife ~ TotalAngle*CuttingSpeed + I(TotalAngle^2) + I(CuttingSpeed^2), data = df2q)
summary(ols2)
```

```{r, eval=TRUE, echo=TRUE}
ols2_full <- lm(ToolLife ~ (TotalAngle + I(TotalAngle^2))*(CuttingSpeed + I(CuttingSpeed^2)), data = df2q)
summary(ols2_full)
```

```{r ols2_full_contour, eval=TRUE, echo=TRUE, fig.align='center', fig.width=6, fig.height=6}
x1_grid <- seq(min(df2q$TotalAngle), max(df2q$TotalAngle), length = 100)
x2_grid <- seq(min(df2q$CuttingSpeed), max(df2q$CuttingSpeed), length = 100)
newx <- data.frame(
  TotalAngle = rep(x1_grid, each = 100),
  CuttingSpeed = rep(x2_grid, time = 100)
)
rs <- predict(ols2_full, newx)

rsplot_data <- data.frame(newx, rs = rs)
rsplot_data$TotalAngle <- rsplot_data$TotalAngle + x1m
rsplot_data$CuttingSpeed <- rsplot_data$CuttingSpeed + x2m

library(ggplot2)
#- Add color to the contour plot
ggplot(rsplot_data) +
  geom_contour(aes(TotalAngle, CuttingSpeed, z = rs), colour = "white") +
  geom_contour_filled(aes(TotalAngle, CuttingSpeed, z = rs))  

```


```{r ols2_full_plotly, eval=TRUE, echo=TRUE, fig.align='center', fig.width=8, fig.height=8}
library(plotly)
rsplot_matrix <- matrix(rsplot_data$rs, 100, 100)
p <- plot_ly(z = rsplot_matrix, type = "surface") %>%
  layout(scene = list(
      xaxis = list(
        title = 'TotalAngle',
        ticktext = lapply(seq(0, 100, 20), function(i) diff(range(rsplot_data$TotalAngle))*i/100 + min(rsplot_data$TotalAngle)), 
        tickvals = list(0, 20, 40, 60, 80, 100),
        tickmode = "array"
      ),
      yaxis = list(
        title = 'CuttingSpeed',
        ticktext = lapply(seq(0, 100, 20), function(i) diff(range(rsplot_data$CuttingSpeed))*i/100 + min(rsplot_data$CuttingSpeed)), 
        tickvals = list(0, 20, 40, 60, 80, 100),
        tickmode = "array"        
      ),
      zaxis = list(title = 'hat(ToolLife)')))

htmlwidgets::saveWidget(as_widget(p), "plotly_rsm_ch5.html")

```




```{r, eval=TRUE, echo=TRUE}
df3 <- read.csv(file.path("data", "5_Impurity.csv"))
df3$Temperature <- as.factor(df3$Temperature)
df3$Pressure <- as.factor(df3$Pressure)
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
fit3 <- aov(Impurity ~ Temperature * Pressure, data = df3)
summary(fit3)
```







```{r, eval=TRUE, echo=TRUE}
df4 <- read.csv(file.path("data", "5_SoftDrinkBottling.csv"))
df4$Carbonation <- as.factor(df4$Carbonation)
df4$Pressure <- as.factor(df4$Pressure)
df4$LineSpeed <- as.factor(df4$LineSpeed)
```



```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
fit4 <- aov(FillHeightsDev ~ Carbonation * Pressure * LineSpeed, data = df4)
summary(fit4)
```



# Chapter 6
 
```{r, eval=TRUE, echo=TRUE}
df5 <- read.csv(file.path("data", "6_ChemicalRecovery.csv"))
```
 
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
# Effect of factor A (reactant concentration)
mean(df5$Recovery[df5$ReactConc == 1] - df5$Recovery[df5$ReactConc == -1])
# Effect of factor A (catalyst amount)
mean(df5$Recovery[df5$CataAmo == 1] - df5$Recovery[df5$CataAmo == -1])
# Effect of two-factor interaction AB
twofi <- as.numeric(df5$ReactConc)*df5$CataAmo
mean(df5$Recovery[twofi == 1] - df5$Recovery[twofi == -1])
```
```{r, eval=TRUE, echo=FALSE, fig.align='center', fig.width=4}
# Effect of factor A (reactant concentration)
cat(sprintf("Effect of factor A (reactant concentration): %.4f\n\n", 
  mean(df5$Recovery[df5$ReactConc == 1] - df5$Recovery[df5$ReactConc == -1])
))
# Effect of factor A (catalyst amount)
cat(sprintf("Effect of factor A (catalyst amount): %.4f\n\n", 
  mean(df5$Recovery[df5$CataAmo == 1] - df5$Recovery[df5$CataAmo == -1])
))
# Effect of two-factor interaction AB
twofi <- as.numeric(df5$ReactConc)*df5$CataAmo
cat(sprintf("Effect of two-factor interaction AB: %.4f\n\n", 
  mean(df5$Recovery[twofi == 1] - df5$Recovery[twofi == -1])
))
```




```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
fit5 <- aov(Recovery ~ factor(ReactConc)*factor(CataAmo), data = df5)
summary(fit5)
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=6, fig.height=6}
interaction.plot(
  x.factor = df5$ReactConc, # x-axis variable
  trace.factor = df5$CataAmo, # variable for lines
  response = df5$Recovery, # y-axis variable
  ylab = "Recovery", xlab = "Reactant Concentration", 
  col = c("firebrick", "dodgerblue2"),
  lty = 1, lwd = 3, trace.label = "Catalyst\nAmount"
)
```



```{r, eval=TRUE, echo=TRUE}
df6 <- read.csv(file.path("data", "6_PlasmaEtching_2^3.csv"))
```



```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
fit6 <- aov(EachRate ~ factor(Gap) * factor(GasFlow) * factor(Power), data = df6)
summary(fit6)
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
ols6_full <- lm(EachRate ~ Gap * GasFlow * Power, data = df6)
summary(ols6_full)
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
ols6_reduced <- lm(EachRate ~ Gap * Power, data = df6)
summary(ols6_reduced)
```


```{r, eval=TRUE, echo=TRUE}
df7 <- read.csv(file.path("data", "6_PilotPlant.csv"))
df7$Temperature <- as.factor(df7$Temperature)
df7$Pressure <- as.factor(df7$Pressure)
df7$FormaldehydeConc <- as.factor(df7$FormaldehydeConc)
df7$StirringRate <- as.factor(df7$StirringRate)
```


```{r halfqqnorm, eval=TRUE, echo=TRUE}
# Half Normal Plot
halfqqnorm <- function(input, tol = 0.5) {
	#input <- eff641
	y <- sort(abs(input))
	nq <- qnorm(seq(0.5, 0.99, length = length(y)))
	plot(y, nq, yaxt = "n", pch = 15, xlab = "|Effects|", ylab = "Half Normal Probability")
	title("Half Normal Plot")
	s <- min(which(diff(y)/diff(range(y)) > 1/(length(y)-1)))
	abline(a = -y[1]*(nq[s]-nq[1])/(y[s]-y[1]), b = (nq[s]-nq[1])/(y[s]-y[1]))
	axis(2, at = qnorm(seq(0.5, 0.9999, length = 5)), labels = round(seq(0, 1, length = 5),2))
	loc <- sqrt((nq - (y - y[1])*(nq[s]-nq[1])/(y[s]-y[1]))^2) > tol
	if (is.null(names(y))) {
		text(y[loc], nq[loc], order(abs(input))[loc], pos = 2)
	} else {
		text(y[loc], nq[loc], names(abs(input))[order(abs(input))[loc]], pos = 2)
	}
}
```


```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
fit7 <- aov(FiltrationRate ~ Temperature * Pressure * FormaldehydeConc * StirringRate, data = df7)
summary(fit7)
```




```{r, eval=TRUE, echo=TRUE}
df8 <- read.csv(file.path("data", "6_Drilling.csv"))
```





```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
fit8 <- aov(AdvRate ~ factor(FlowRate) * (factor(RotatSp) + factor(MudType)), data = df8)
summary(fit8)
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4, fig.height=4}
interaction.plot(
  x.factor = df8$FlowRate, # x-axis variable
  trace.factor = df8$RotatSp, # variable for lines
  response = df8$AdvRate, # y-axis variable
  ylab = "AdvRate", xlab = "FlowRate", 
  col = c("firebrick", "dodgerblue2"),
  lty = 1, lwd = 3, trace.label = "RotatSp"
)

interaction.plot(
  x.factor = df8$FlowRate, # x-axis variable
  trace.factor = df8$MudType, # variable for lines
  response = df8$AdvRate, # y-axis variable
  ylab = "AdvRate", xlab = "FlowRate", 
  col = c("firebrick", "dodgerblue2"),
  lty = 1, lwd = 3, trace.label = "MudType"
)
```


```{r, eval=TRUE, echo=TRUE}
df9 <- read.csv(file.path("data", "6_Sidewall.csv"))
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4}
fit9 <- aov(AvgDefectNum ~ factor(Temperature) + factor(ResinFlow), data = df9)
summary(fit9)
```


```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=6, fig.height=6}
par(mfrow = c(2, 2))
plot(fit9)
par(mfrow = c(1, 1))
```


```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=4, fig.height=4}
plot(
  as.numeric(df9$ClampTime), fit9$residuals, pch = 16,
  xlab = "ClampTime", ylab = "residuals", xlim = c(-2, 2)
)
abline(h = 0, lty = 3)
```




```{r, eval=TRUE, echo=TRUE}
df10 <- read.csv(file.path("data", "6_Wafer.csv"))
```

```{r, eval=TRUE, echo=TRUE}
df11 <- read.csv(file.path("data", "6_CreditCardMarketing.csv"))
```



