---
title: "Ch-06 R Codes"
author: "Ping-Yang Chen"
#date: "2024-03-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Textbook: Montgomery, D. C. (2012). *Design and analysis of experiments*, 8th Edition. John Wiley & Sons.

Online handouts: https://github.com/PingYangChen/ANOVA_Course_R_Code

## $2^2$-Design: Chemical Process Example
 
Read the csv file `6_ChemicalRecovery.csv` in R.
```{r, eval=TRUE, echo=TRUE}
df5 <- read.csv(file.path("data", "6_ChemicalRecovery.csv"))
```


The estimation of factor effects are:
```{r, eval=FALSE, echo=TRUE, fig.align='center', fig.width=5}
# Effect of factor A (reactant concentration)
mean(df5$Recovery[df5$ReactConc == 1] - df5$Recovery[df5$ReactConc == -1])
# or 2*mean(df5$Recovery*df5$ReactConc)

# Effect of factor A (catalyst amount)
mean(df5$Recovery[df5$CataAmo == 1] - df5$Recovery[df5$CataAmo == -1])

# Effect of two-factor interaction AB
twofi <- as.numeric(df5$ReactConc)*df5$CataAmo
mean(df5$Recovery[twofi == 1] - df5$Recovery[twofi == -1])
```
```{r, eval=TRUE, echo=FALSE, fig.align='center', fig.width=5}
# Effect of factor A (reactant concentration)
cat(sprintf("Effect of factor A (reactant concentration): %.4f\n\n", 
  mean(df5$Recovery[df5$ReactConc == 1] - df5$Recovery[df5$ReactConc == -1])
))
# Effect of factor A (catalyst amount)
cat(sprintf("Effect of factor A (catalyst amount): %.4f\n\n", 
  mean(df5$Recovery[df5$CataAmo == 1] - df5$Recovery[df5$CataAmo == -1])
))
# Effect of two-factor interaction AB
twofi <- as.numeric(df5$ReactConc)*df5$CataAmo
cat(sprintf("Effect of two-factor interaction AB: %.4f\n\n", 
  mean(df5$Recovery[twofi == 1] - df5$Recovery[twofi == -1])
))
```


Use `aov()` to fit the ANOVA model (model description in math is omitted).
Here, if the columns of the factor in the `data.frame` are not of the `factor` type. We can also specify the type of the factors as `factor` in the R model formula.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit5 <- aov(Recovery ~ factor(ReactConc)*factor(CataAmo), data = df5)
summary(fit5)
```

The two main effects are significant and their two-factor interaction may not exist.

To visualize the analysis result of the factorial experiment, interaction plot is commonly used tool.
```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
interaction.plot(
  x.factor = df5$ReactConc, # x-axis variable
  trace.factor = df5$CataAmo, # variable for lines
  response = df5$Recovery, # y-axis variable
  ylab = "Recovery", xlab = "Reactant Concentration", 
  col = c("firebrick", "dodgerblue2"),
  lty = 1, lwd = 3, trace.label = "Catalyst\nAmount"
)
```


## $2^3$-Design: Plasma Etching Example

Read the csv file `6_PlasmaEtching_2^3.csv` in R.
```{r, eval=TRUE, echo=TRUE}
df6 <- read.csv(file.path("data", "6_PlasmaEtching_2^3.csv"))
```

The estimation of factor effects are:
```{r, eval=FALSE, echo=TRUE, fig.align='center', fig.width=5}
# Effect of main effects
2*mean(df6$EachRate*df6$Gap)
2*mean(df6$EachRate*df6$GasFlow)
2*mean(df6$EachRate*df6$Power)

# Effect of two-factor interactions
GapGasFlow <- df6$Gap * df6$GasFlow
GapPower <- df6$Gap * df6$Power
GasFlowPower <- df6$GasFlow * df6$Power
2*mean(df6$EachRate*GapGasFlow)
2*mean(df6$EachRate*GapPower)
2*mean(df6$EachRate*GasFlowPower)

# Effect of three-factor interaction
GapGasFlowPower <- df6$Gap * df6$GasFlow * df6$Power
2*mean(df6$EachRate*GapGasFlowPower)
```

```{r df6eff, eval=TRUE, echo=FALSE, fig.align='center', fig.width=5}
# Effect of two-factor interactions
GapGasFlow <- df6$Gap * df6$GasFlow
GapPower <- df6$Gap * df6$Power
GasFlowPower <- df6$GasFlow * df6$Power
# Effect of two-factor interaction AB
GapGasFlowPower <- df6$Gap * df6$GasFlow * df6$Power

effs <- c(
  2*mean(df6$EachRate*df6$Gap),
  2*mean(df6$EachRate*df6$GasFlow),
  2*mean(df6$EachRate*df6$Power),
  2*mean(df6$EachRate*GapGasFlow),
  2*mean(df6$EachRate*GapPower),
  2*mean(df6$EachRate*GasFlowPower),
  2*mean(df6$EachRate*GapGasFlowPower)
)

eff6 <- data.frame(
  Factor = c("A", "B", "C", "AB", "AC", "BC", "ABC"),
  "Est.Effect" = effs
)
print(eff6)
```

Use `aov()` to fit the ANOVA model (model description in math is omitted).
Here, if the columns of the factor in the `data.frame` are not of the `factor` type. We can also specify the type of the factors as `factor` in the R model formula.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit6 <- aov(
  EachRate ~ factor(Gap) * factor(GasFlow) * factor(Power), 
  data = df6
)
summary(fit6)
```

The significant effects are main effects `Gap` and `Power`, and, the two-factor interaction `Gap:Power`.

WE can also fit a response surface model and obtain the same conslusion.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
ols6_full <- lm(EachRate ~ Gap * GasFlow * Power, data = df6)
summary(ols6_full)
```


The final RSM is the reduced model that those insignificant terms are removed from the full model. Now we can use the real value of the factot levels as the predictors' values fro this reduced model for better interpretation.
```{r ols6_reduced, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
ols6_reduced <- lm(EachRate ~ NumGap * NumPower, data = df6)
summary(ols6_reduced)
```

```{r ols6_reduced_contour, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
x1_grid <- seq(min(df6$NumGap), max(df6$NumGap), length = 100)
x2_grid <- seq(min(df6$NumPower), max(df6$NumPower), length = 100)
newx <- data.frame(
  NumGap = rep(x1_grid, each = 100),
  NumPower = rep(x2_grid, time = 100)
)

rsplot_data6 <- data.frame(newx, rs = predict(ols6_reduced, newx))

library(ggplot2)
#- Add color to the contour plot
ggplot(rsplot_data6) +
  geom_contour(aes(NumGap, NumPower, z = rs), colour = "white") +
  geom_contour_filled(aes(NumGap, NumPower, z = rs))  
```

## Unreplicated $2^k$ Factorial Designs: The Resin Plant Experiment

Read the csv file `6_PilotPlant.csv` in R.
```{r, eval=TRUE, echo=TRUE}
df7 <- read.csv(file.path("data", "6_PilotPlant.csv"))
```

The estimation of factor effects are:
```{r eff7, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
# Compute the model matrix of all effect terms without intercept
mmat7 <- model.matrix( ~ Temperature*Pressure*CH2OConc*StirRate - 1, data = df7)
# Calculate the effect sizes using the +/- signs of the model matrix
eff7 <- numeric(ncol(mmat7))
for (i in 1:ncol(mmat7)) {
  eff7[i] <- 2*mean(df7$FiltrationRate*mmat7[,i])
}
names(eff7) <- colnames(mmat7)
```

```{r, eval=TRUE, echo=FALSE, fig.align='center', fig.width=5}
print(data.frame(
  Factor = names(eff7),
  "Est.Effect" = as.vector(eff7)
))
```

Because there is no replicates in each treatment combination, the estimate of the random error $\sigma^2$ does not exist and the ANOVA table is not available. Of all effect terms, we try of eliminate some of them before analyzing the data. According to the model assumption, the effects that are negligible should be similar to random error which is normally distributed with zero mean and constant variance. Therefore, a QQ-plot or half-Normal plot is helpful to identify effective effects. Belows are the codes of half-Normal plot.

```{r halfqqnorm, eval=TRUE, echo=TRUE}
# Half Normal Plot
halfqqnorm <- function(input, tol = 0.5) {
	y <- sort(abs(input))
	nq <- qnorm(seq(0.5, 0.99, length = length(y)))
	plot(y, nq, yaxt = "n", pch = 15, 
	     xlab = "|Effects|", ylab = "Half Normal Probability")
	title("Half Normal Plot")
	# choose anchor point to draw a straight line
	s <- min(which(diff(y)/diff(range(y)) > 1/(length(y)-1)))
	abline(a = -y[1]*(nq[s]-nq[1])/(y[s]-y[1]), b = (nq[s]-nq[1])/(y[s]-y[1]))
	axis(2, at = qnorm(seq(0.5, 0.9999, length = 5)), 
	     labels = round(seq(0, 1, length = 5),2))
	loc <- sqrt((nq - (y - y[1])*(nq[s]-nq[1])/(y[s]-y[1]))^2) > tol
	if (is.null(names(y))) {
		text(y[loc], nq[loc], order(abs(input))[loc], pos = 2)
	} else {
		text(y[loc], nq[loc], names(abs(input))[order(abs(input))[loc]], pos = 2)
	}
}
```


```{r h7, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
halfqqnorm(eff7)
```

By the half Normal plot, we find out that the main effects `Temperature`, `CH2OConc`and `StirRate` and interactions `Temperature:CH2OConc`, `Temperature:StirRate` appear to be large.

Based on the observation from the half Normal plot, now the ANOVA model is

\begin{equation}
y_{ijkl} = \mu + \tau_i + \beta_j + \gamma_k + (\tau\beta)_{ij} + (\tau\gamma)_{ik} + \varepsilon_{ijkl}
\end{equation}

- $\tau_i$ is the effect of the $i$th `Temperature` level, $i = 1, 2$.
- $\beta_j$ is the effect of the $j$th `CH2OConc` level, $j = 1, 2$.
- $\gamma_k$ is the effect of the $k$th `StirRate` level, $j = 1, 2$.
- $(\tau\beta)_{ij}$ is the interaction effect of the $i$th `Temperature` level and the $j$th `CH2OConc` level.
- $(\tau\gamma)_{ij}$ is the interaction effect of the $i$th `Temperature` level and the $k$th `StirRate` level.
- $\varepsilon_{ijkl}$ is the random error, $l = 1, 2$, satisfying
$$\varepsilon_{ijkl}\overset{i.i.d.}{\sim}N(0, \sigma^2) \text{ where } \sigma^2 \text{ is the conatnt variance.}$$

Use `aov()` to fit the ANOVA model.

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit7 <- aov(
  FiltrationRate ~ factor(Temperature) * (factor(CH2OConc) + factor(StirRate)), 
  data = df7
)
summary(fit7)
```

Here, we left the interpretation of the response surface model for practice.

<!--
```{r, eval=TRUE, echo=TRUE}
df8 <- read.csv(file.path("data", "6_Drilling.csv"))
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit8 <- aov(AdvRate ~ factor(FlowRate) * (factor(RotatSp) + factor(MudType)), data = df8)
summary(fit8)
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
interaction.plot(
  x.factor = df8$FlowRate, # x-axis variable
  trace.factor = df8$RotatSp, # variable for lines
  response = df8$AdvRate, # y-axis variable
  ylab = "AdvRate", xlab = "FlowRate", 
  col = c("firebrick", "dodgerblue2"),
  lty = 1, lwd = 3, trace.label = "RotatSp"
)

interaction.plot(
  x.factor = df8$FlowRate, # x-axis variable
  trace.factor = df8$MudType, # variable for lines
  response = df8$AdvRate, # y-axis variable
  ylab = "AdvRate", xlab = "FlowRate", 
  col = c("firebrick", "dodgerblue2"),
  lty = 1, lwd = 3, trace.label = "MudType"
)
```

```{r, eval=TRUE, echo=TRUE}
df9 <- read.csv(file.path("data", "6_Sidewall.csv"))
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit9 <- aov(AvgDefectNum ~ factor(Temperature) + factor(ResinFlow), data = df9)
summary(fit9)
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=6, fig.height=6}
par(mfrow = c(2, 2))
plot(fit9)
par(mfrow = c(1, 1))
```

```{r, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
plot(
  as.numeric(df9$ClampTime), fit9$residuals, pch = 16,
  xlab = "ClampTime", ylab = "residuals", xlim = c(-2, 2)
)
abline(h = 0, lty = 3)
```

```{r, eval=TRUE, echo=TRUE}
df10 <- read.csv(file.path("data", "6_Wafer.csv"))
```

```{r, eval=TRUE, echo=TRUE}
df11 <- read.csv(file.path("data", "6_CreditCardMarketing.csv"))
```
-->

## Addition of Center Points to a $2^k$ Designs 

Recall the Resin Plant Experiment. Read the csv file `6_PilotPlant.csv` in R.

```{r df7a, eval=TRUE, echo=TRUE}
df7 <- read.csv(file.path("data", "6_PilotPlant.csv"))
```

```{r df7c, eval=TRUE, echo=TRUE}
df7_C <- data.frame(
  Temperature = rep(0, 4),
  Pressure = rep(0, 4),
  CH2OConc = rep(0, 4),
  StirRate = rep(0, 4),
  FiltrationRate = c(73, 75, 66, 69)
)
```

The calculation of $SS_{Pure quadratic}$ with degree of freedom 1.
```{r sspq, eval=TRUE, echo=TRUE}
Yf_bar <- mean(df7$FiltrationRate)
Yc_bar <- mean(df7_C$FiltrationRate)
nf <- nrow(df7)
nc <- nrow(df7_C)
SS_pureQuad <- nf*nc*(Yf_bar - Yc_bar)^2/(nf + nc)
```

The calculation of $SS_{E}$ with degree of freedom $n_c - 1$.

```{r sse, eval=TRUE, echo=TRUE}
SS_E <- sum((df7_C$FiltrationRate - mean(df7_C$FiltrationRate))^2)
```

To test the significance of the Curvature, we compute the ratio of $MS_{Pure quadratic}$ and $MS_{E}$ as the $F$-statistic which follows the $F$ distribution with degrees of freedom 1 and $n_c - 1$.
```{r, eval=TRUE, echo=TRUE}
MS_pureQuad <- SS_pureQuad/1
MS_E <- SS_E/(nc - 1)
fval <- MS_pureQuad/MS_E
pval <- 1 - pf(fval, 1, nc - 1) # p-value
```

The part of the testing the significance of the Curvature of the ANOVA table.
```{r, eval=TRUE, echo=TRUE}
print(data.frame(
  Source = c("Curvature", "Residual"),
  SS = c(SS_pureQuad, SS_E),
  DF = c(1, nc - 1), 
  MS = c(MS_pureQuad, MS_E),
  "F" = c(sprintf("%.3f", fval), ""),
  "Pr(>F)" = c(sprintf("%.3f", pval), "")
))
```
