---
title: "Ch-07_08 R Codes"
author: "Ping-Yang Chen"
date: "2024-05-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Textbook: Montgomery, D. C. (2012). *Design and analysis of experiments*, 8th Edition. John Wiley & Sons.

Online handouts: https://github.com/PingYangChen/ANOVA_Course_R_Code

# 7.21

By the defining contrast, to confound eight blocks with $ABCD$, $ACE$ and $ABEF$, let

\begin{align*}
L_1 & = x_1 + x_2 + x_3 + x_4 \\
L_2 & = x_1 + x_3 + x_5 \\
L_3 & = x_1 + x_2 + x_5 + x_6 \\
\end{align*}



```{r r721, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
designMat <- data.frame(
	A = rep(0:1, 32),
	B = rep(rep(0:1, each = 2), 16),
	C = rep(rep(0:1, each = 4), 8),
	D = rep(rep(0:1, each = 8), 4),
	E = rep(rep(0:1, each = 16), 2),
	F = rep(0:1, each = 32)
)
print(head(designMat, 6))

letterMat <- sapply(1:ncol(designMat), function(j) {
  ifelse(designMat[,j] == 1, letters[j], "")  
})
print(head(letterMat, 6))

effectNames <- sapply(1:nrow(letterMat), function(i) {
  ifelse(all(letterMat[i,] == ""), "(1) ", paste0(letterMat[i,], collapse = ""))
})
rownames(designMat) <- effectNames
print(head(designMat, 6))
```

```{r r721-1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
attach(designMat)
assignBlock <- data.frame(
	L1 = (A + B + C + D) %% 2,
	L2 = (A + C + E) %% 2,
	L3 = (A + B + E + F) %% 2
)
detach(designMat)
```

```{r r721-2, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
blockId <- as.matrix(assignBlock) %*% c(2^2, 2, 1) + 1
```



```{r r721-3, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
result <- matrix("", 8, 8)
for (i in 1:8) {
  result[,i] <- effectNames[which(blockId == i)]
}
colnames(result) <- sprintf("Block %d", 1:8)
print(data.frame(result))
```
The other effects confounded with blocks:


\begin{align*}
(ABCD)(ACE) & = A^2 B C^2 D E = BDE \\
(ABCD)(ABEF) & = A^2 B^2 C D E F = CDEF \\
(ACE)(ABEF) & = A^2 B C E^2 F = BCF \\
(ABCD)(BCF) & = A B^2 C^2 D F = ADF 
\end{align*}


# 8.11

$2^{5-2}$ fractional factorial design with defining relation $$I = ACE \text{ and } I = BDE$$

1. generate the $2^3$ full factorial design

```{r r811, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
lvl <- c(-1, 1)
FF3 <- data.frame(
	A = rep(lvl, 4),
	B = rep(rep(lvl, each = 2), 2),
	C = rep(lvl, each = 4)
)
print(FF3)
```


2. add two columns $D$ and $E$ to form the $2^{5-2}$ by the defining relation 

\begin{align*}
I = ACE & \implies E = AC \\
I = BDE & \implies D = BE \implies D = BAC \\
\end{align*}

```{r r811-1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
attach(FF3)
augmentFrF <- data.frame(
	D = A * B * C,
	E = A * C
)
detach(FF3)

FrF5_2 <- cbind(FF3, augmentFrF)
# Get letters of each effect
letterMat <- sapply(1:ncol(FrF5_2), function(j) {
  ifelse(FrF5_2[,j] == 1, letters[j], "")  
})
# Combine letters 
effectNames <- sapply(1:nrow(letterMat), function(i) {
  ifelse(all(letterMat[i,] == ""), "(1) ", paste0(letterMat[i,], collapse = ""))
})
rownames(FrF5_2) <- effectNames
print(FrF5_2)
```


Complete defining relation is

$$I = ACE = BDE = ABCD$$

All aliases are

\begin{align*}
A & = CE = BCD \\
B & = DE = ACD \\ 
C & = AE = ABD \\
D & = BE = ABC \\
E & = AC \\
AB & = CD \\
AD & = BC \\
AC & = BD \\
\end{align*}



```{r r811-2, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
y <- numeric(8)
y[effectNames == "e"]     <- 23.2
y[effectNames == "ad"]    <- 16.9
y[effectNames == "cd"]    <- 23.8
y[effectNames == "bde"]   <- 16.8
y[effectNames == "ab"]    <- 15.5
y[effectNames == "bc"]    <- 16.2
y[effectNames == "ace"]   <- 23.4
y[effectNames == "abcde"] <- 18.1
frfData <- cbind(FrF5_2, y = y)
```

The estimation of factor effects are:
```{r r811-3, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
# Compute the model matrix of all effect terms without intercept
mmat5 <- model.matrix( ~ A+B+C+D+E+A:B+A:D - 1, data = frfData)
# Calculate the effect sizes using the +/- signs of the model matrix
eff5 <- numeric(ncol(mmat5))
for (i in 1:ncol(mmat5)) {
  eff5[i] <- 2*mean(frfData$y*mmat5[,i])
}
names(eff5) <- colnames(mmat5)
```

```{r, eval=TRUE, echo=FALSE, fig.align='center', fig.width=5}
print(data.frame(
  Factor = names(eff5),
  "Est.Effect" = as.vector(eff5)
))
```


```{r halfqqnorm, eval=TRUE, echo=TRUE}
# Half Normal Plot
halfqqnorm <- function(input, tol = 0.5) {
	y <- sort(abs(input))
	nq <- qnorm(seq(0.5, 0.99, length = length(y)))
	plot(y, nq, yaxt = "n", pch = 15, 
	     xlab = "|Effects|", ylab = "Half Normal Probability")
	title("Half Normal Plot")
	# choose anchor point to draw a straight line
	s <- min(which(diff(y)/diff(range(y)) > 1/(length(y)-1)))
	abline(a = -y[1]*(nq[s]-nq[1])/(y[s]-y[1]), b = (nq[s]-nq[1])/(y[s]-y[1]))
	axis(2, at = qnorm(seq(0.5, 0.9999, length = 5)), 
	     labels = round(seq(0, 1, length = 5),2))
	loc <- sqrt((nq - (y - y[1])*(nq[s]-nq[1])/(y[s]-y[1]))^2) > tol
	if (is.null(names(y))) {
		text(y[loc], nq[loc], order(abs(input))[loc], pos = 2)
	} else {
		text(y[loc], nq[loc], names(abs(input))[order(abs(input))[loc]], pos = 2)
	}
}
```




```{r h5, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5, fig.height=5}
halfqqnorm(eff5)
```


```{r r811-4, eval=TRUE, echo=TRUE, fig.align='center', fig.width=5}
fit2 <- aov(y ~ B, data = frfData)
summary(fit2)
```
